---
title: "Tidyverse Core 1: filtering, selecting and arranging data"
---

```{r echo=FALSE}

knitr::opts_chunk$set(class.source='fold-show')

```

#### Learning Objectives
In this lesson, you will learn how to:

- `filter()` out only the observations that satisfy a specific criteria from the dataset
- `select()`  the columns we're interested in
- `arrange()` the output in a specific order
- Combine multiple filtering and selecting steps using the "pipe" `%>%` operator

#### Questions

- How can I filter rows and/or columns from my data frame based on specific criteria?
- How can I chain together multiple filtering and selecting statements to return only the rows and columns I'm interested in?
- How can I arrange my results in descending or ascending order?

***

## Filtering data

Now that we've had a look at our dataset, let's learn the first of the tidyverse's core commands: `dplyr::filter()`. This commands filters our rows in the dataset that match a specific criteria. This is similar to Excel's filter function, where you can check off which values you'd like to see the rows for, or which values are greater than or less than a specific cut point. 

Let's use the `glimpse()` command, introduced in the last lesson, to remind ourselves about the variables in the gapminder dataset we're working with:

```{r loadLibraries, message=FALSE}
library(tidyverse)
library(gapminder)
glimpse(gapminder)
```

Next, let's use the `table()` command to see how many records for each year we have:

```{r years}
table(gapminder$year)
```

Let's filter out the records from 1952 using the `filter()` command:

```{r filter1}
filter(gapminder, year == 1952)
```

If we are interested in retrieving records from the 1950s:

```{r filter2}
filter(gapminder, year < 1960)
```
Notice that 284 rows are returned instead of 142, as in the previous example. 

We can also combine `filter()` statements. 

For example, to filter out records from Oceania from the 1950s we can do:

```{r filter3}
filter(gapminder, year < 1960 & continent == "Oceania")
```

This returns only four records, one each for Australia and NZ in 1952 and 1957.

The `|` operator can be used to specify "or", for example to filter out records from the 1950s from Australia and New Zealand explicitly:

```{r filter4}
filter(gapminder, year < 1960 & (country == "Australia" | country == "New Zealand"))
```

If there are more than two items you would like to consider, you can use a character vector and the `%in%` command to filter instead:

```{r filter5}
mycharactervector <- c("Australia", "New Zealand", "Argentina")

filter(gapminder, year < 1960 & country %in% mycharactervector)
```

We will use the `%in%` command and character vectors more in later sessions.

### Challenge 1

Which of these commands would return records for which the life expectancy was between 50 and 59 in 1992?


- A: `filter(gapminder, lifeExp >= 50 & lifeExp <= 60 & year == 1992)`
- B: `filter(gapminder, lifeExp >= 50 | lifeExp < 60 | year == 1992)`
- C: `filter(gapminder, lifeExp >= 50 & lifeExp <= 59 & year == 1992)`
- D: `filter(gapminder, lifeExp >= 50 | (lifeExp < 60 & year == 1992))`
- E: `filter(gapminder, (lifeExp >= 50 & lifeExp <= 60) & year == 1992)`

```{r Solution1, class.source = NULL, eval =FALSE}
# Correct answer: C
filter(gapminder, lifeExp >= 50 & lifeExp <= 59 & year == 1992)
# A: also includes countries with a LE of 60 years. Testing "<=" handling
filter(gapminder, lifeExp >= 50 & lifeExp <= 60 & year == 1992)
# B: testing for an understanding of how "or" works - returns entire dataset.
filter(gapminder, lifeExp >= 50 | lifeExp < 60 | year == 1992)
# D: testing for an understanding of how brackets work 
filter(gapminder, lifeExp >= 50 | (lifeExp < 60 & year == 1992))
# E: testing for an understanding of how brackets and edge cases work
filter(gapminder, (lifeExp >= 50 & lifeExp <= 60) & year == 1992)

```

### Challenge 2

Use the `filter()` command to extract records from the Americas from the 1970s.


```{r Solution2, class.source = NULL, eval =FALSE}
# We can combine more than two statements in our filtering criteria:
filter(gapminder, year >= 1970 & year < 1980 & continent == "Americas")
#
# We can also use dplyr::between() 
filter(gapminder, between(year, 1970,1979) & continent == "Americas")
#
# Watch out, as between included both the left and right hand side in the filtering criteria,
# So records from both 1970 and 1979 are returned with the above statement!
```

### Challenge 3

Which of these (more than one is correct) would return records from Argentina and Australia from the 1980s? Which one of the correct solutions do you think is the most readable?

- A: `filter(gapminder, (year < 1989 & year > 1980) & (country == "Argentina" | country == "Australia"))` 
- B: `filter(gapminder, (year < 1989 & country == "Argentina") & (year > 1980 | country == "Australia"))`
- C: `filter(gapminder, (year <= 1989 & year >= 1980) & (country == "Argentina" | country == "Australia"))` 
- D: `filter(gapminder, year <= 1989 & year >= 1980 & country == "Argentina" | country == "Australia")` 
- E: `filter(gapminder, (year < 1990 & year > 1979) & (country == "Argentina" | country == "Australia"))` 
- F: `filter(gapminder, (year <= 1989 & year >= 1980 & country == "Argentina" ) | (year <= 1989 & year >= 1980 &  country == "Australia"))`
- G: `filter(gapminder, (year == 1982 | year == 1987) & (country == "Argentina" | country == "Australia"))` 
- H: `filter(gapminder, year <= 1989 | year >= 1980 | country == "Argentina" | country == "Australia")` 

```{r Solution3, class.source = NULL, eval =FALSE}
# The following three (C, E, F) are *the* correct solutions, with the first one probably being the most readable, as 
# it makes explicit the fact that we're pulling out records from the 1980s:
filter(gapminder, (year <= 1989 & year >= 1980) & (country == "Argentina" | country == "Australia"))
filter(gapminder, (year < 1990 & year > 1979) & (country == "Argentina" | country == "Australia"))
filter(gapminder, (year <= 1989 & year >= 1980 & country == "Argentina" ) | (year <= 1989 & year >= 1980 &  country == "Australia"))


# The following two (A, G) also producte the correct result (so are correct in this instance), but would not 
# generalise to other dataset, or to our dataset were we to include more data, say from 1989
filter(gapminder, (year < 1989 & year > 1980) & (country == "Argentina" | country == "Australia"))
filter(gapminder, (year == 1982 | year == 1987) & (country == "Argentina" | country == "Australia"))

# Finally, the following approaches (B, D, H) do not generate the desired results at all:
# this returns all records from Argentina between 1981 and 1988.
filter(gapminder, (year < 1989 & country == "Argentina") & (year > 1980 | country == "Australia"))
# this returns all the records from 1980 to 1989 in Argentina and all of the records (for all time) from Australia
filter(gapminder, year <= 1989 & year >= 1980 & country == "Argentina" | country == "Australia")
# this actually does nothing as it returns all records which are before 1989 OR after 1980, so all records
# in the dataset
# the country filtering does nothing at all
filter(gapminder, year <= 1989 | year >= 1980 | country == "Argentina" | country == "Australia")

```



***

## Using pipes to combine multiple dplyr operations

While the filtering commands we just implemented were not very long and relatively straightforward to read, often when we work with real-world datasets we can have quite complex filtering strategies. For example, if we wanted to filter records from the 1970s from countries with a population greater than 10 million in the Americas with a life expectancy greater than 65 we could write: 

```{r veryLongFilter}
filter(gapminder, continent == "Americas" & year < 1980 & year >= 1970 & lifeExp > 65 & pop > 10000000) 
```

However, this becomes quite cumbersome to read. Instead, we can use the so-called pipe operator, `%>%`, to chain together commands. The pipe operator is named this way because it works something like a water pipe - when we put a filter on our water pipes, some of the stuff gets caught, but most of the water flows through. Similarly, in programming, the output of the previous command is piped in straight though as input to the next command. So, to rewrite the convoluted statement above, we could do the following:

```{r pipe}
# take our entire gapminder dataset & feed it as input into the first filter command
gapminder %>%
  # filter out all countries that are not in the Americas
  filter(continent == "Americas") %>%
  # filter out all records that are not from the 1970s
  filter(year >= 1970 & year < 1980) %>%
  # filter based on life expectancy and population
  filter(lifeExp > 65 & pop > 10^7)

```

While I have included comments in the code above, they are not essential for deciphering what the code does at each step:

```
gapminder %>%
  filter(continent == "Americas") %>%
  filter(year >= 1970 & year < 1980) %>%
  filter(lifeExp > 65 & pop > 10^7)
``` 

Another advantage of using pipes is that RStudio will help us with tab completion and suggest the names of columns in our data when we use them. This is why we will continue to use the pipe for the rest of this training course.


***
## Arranging our records using the arrange command

We can arrange our data by using the `arrange()` command. For example, to arrange the gapminder dataset by population:

```{r arrange1}
gapminder %>%
  arrange(pop)
```

We can arrange by multiple columns sequentially by listing them in the order we'd like to arrange. For example, to sort by country then population:

```{r arrange1a}
gapminder %>%
  arrange(country, pop)
```


By default, `arrange()` arranges things from lowest to highest, but we can use `desc()` to arrange in descending order:

```{r arrange2}
gapminder %>%
  arrange(desc(pop))
```

Thanks to the pipe, we can combine `arrange()` and `filter()` statements:

```{r arrangeFilter}
gapminder %>%
  filter(continent == "Oceania") %>%
  filter(year < 1960) %>%
  arrange(desc(lifeExp))
```

We can see that the life expectancy in New Zealand was higher than in Australia in 1952, but that in 1957 the opposite became true!

### Challenge 4

Which countries had the highest and second-highest life expectancy in Europe in 1952? Were they still the top two countries in 2007?

```{r Solution4, class.source = NULL, eval =FALSE}
# Norway and Iceland had the highest life expectancy in 1952, with Norway having .18 years (approx 2 months) higher 
# Life expectancy than Iceland
gapminder %>%
  filter(continent == "Europe") %>%
  filter(year == 1952) %>%
  arrange(desc(lifeExp))

# In 2007, Iceland had the highest life expectancy, whereas Norway was relegated to 7th place
# The difference between first and 7th in 2007 was 1.561 years!
gapminder %>%
  filter(continent == "Europe") %>%
  filter(year == 2007) %>%
  arrange(desc(lifeExp))

```

***
## Selecting the columns we want using the `select()` command

So far, we've returned all of the columns from the data frames we've been working with. However, we may be interested in working with only a selection of them, so we can use the `select()` command to pick and choose which columns we are interested in.

For example, to `select()` only the country, year and lifeExp columns from our dataset we can do the following:

```{r select1}
gapminder %>%
  select(country, year, lifeExp)
```

We can combine selecting columns with filtering rows, for example by building on our case study above:

```{r select2}
gapminder %>%
  filter(year < 1960 & continent == "Oceania") %>%
  select(country, year, gdpPercap) %>%
  arrange(gdpPercap)

```

We can see that in the 1950s, both Australia and New Zealand's GDP grew between 1952 and 1957, and Australia's GDP was less than that of New Zealand.

### Challenge 5

- Which countries with a population over 1 million people had the highest GDP in Asia in 1952? 
- How many of them were in the top 5 in 2007?

Note: you can use two separate dplyr expressions to answer both of these questions.
Hint: explore countries' GDP without filtering by population in 1952 first to explain what is occuring in the data

```{r challenge5, class.source = NULL, eval =FALSE}

# First, identify the top countries by GDP in Asia in 1952 without filtering by GDP
gapminder %>%
  filter(year == 1952 & continent == "Asia") %>%
  select(country, year, pop, gdpPercap) %>%
  arrange(desc(gdpPercap)) 
# We can see that it is Kuwait, Bahrain, Saudi Arabia, Lebanon, Iraq, Israel and Japan

# Now filter by population
gapminder %>%
  filter(year == 1952 & continent == "Asia" & pop > 10^6) %>%
  select(country, year, pop, gdpPercap) %>%
  arrange(desc(gdpPercap))
# Kuwait and Bahrain are filtered out, as they have very low population

# Next, look in 2007
gapminder %>%
  filter(year == 2007 & continent == "Asia") %>%
  select(country, year, pop, gdpPercap) %>%
  arrange(desc(gdpPercap))

# We can see that in 2007, the order is Kuwait, Singapore, Hong Kong, Japan, Bahrain
# When we then filter by population, Bahrain (but not Kuwait) is once again filtered out
gapminder %>%
  filter(year == 2007 & continent == "Asia" & pop > 10^6) %>%
  select(country, year, pop, gdpPercap) %>%
  arrange(desc(gdpPercap))
```


***

### Key Points

- The dplyr library of the tidyverse family provides us with the key `filter()`, `select()`, and `arrange()` commands 
- `filter()` filters our rows from our dataset to return the observations that satisfy a specific criteria
- `select()`  selects which columns are returned to us from the dataset
- `arrange()` arranges the output in ascending order of a specific variable
- We can use `desc()` to change the way in which this ordering is done
- The "pipe" `%>%` operator allows us to combine multiple dplyr statements into a long chain, which improves readability and gives us access to RStudio code completion








